Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-15 paper/paper.tex
Started section 4; extended one of the lemmas.

% Shorthands
\newcommand{\kmer}[1]{$#1$\nobreakdash-mer}
\newcommand{\kcollection}[1]{$#1$\nobreakdash-collection}
\newcommand{\kequivalent}[1]{$#1$\nobreakdash-equivalent}
\newcommand{\orderk}[1]{order\nobreakdash-$#1$}
\newcommand{\LFmapping}{LF\nobreakdash-mapping}
\newcommand{\FMindex}{FM\nobreakdash-index}

@@

CST-NPR: \cite{Fischer2009a}

CST MEMs: \cite{Ohlebusch2010a}

@@

The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} is a permutation of the text with the same combinatorial structure as the suffix array. Given text $T[0, n-1]$, its Burrows-Wheeler transform is a string $\BWT[0, n-1]$, where $\BWT[i] = T[(\SA[i]-1) \bmod n]$. Given the \emph{lexicographic rank} $i$ of suffix $T[\SA[i], n-1]$, we can use \emph{\LFmapping} on the BWT to find the lexicographic rank of the previous suffix $T[(\SA[i]-1) \bmod n, n-1]$. Let
$$
\LF(i) = \Carray[\BWT[i]] + \BWT.\rank(i, \BWT[i]),
$$
where $\Carray[c]$ is the number of suffixes of the text starting with any character $c' < c$. Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$. We can also generalize the definition for any character $c \in \Sigma$:
$$
\LF(i, c) = \Carray[c] + \BWT.\rank(i, c).
$$
Let $X$ be a string. If there are $i$ suffixes $S'$ of text $T$ such that $S' < X$ in lexicographic order, then there are $\LF(i, c)$ suffixes smaller than string $cX$ in lexicographic order.

@@

\begin{lemma}[No false negatives]\label{lemma:pg-fn}
Let $G' = (V', E')$ be a path graph of graph $G = (V, E)$, and let $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ be a pattern string. Set $G'.\locate(X)$ contains the start nodes of all paths $P$ in graph $G$ with $G.\glabel(P) = X$.
\end{lemma}

@@

\begin{definition}[Equivalent path graphs]
Let $G$ and $G'$ be two path graphs, and let $k > 0$ be a parameter value. We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ with $1 \le \abs{X} \le k$.
\end{definition}

@@

\begin{lemma}[No short false positives]\label{lemma:dbg-fp}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$, and let $P' = v'_{0} \dotsm v'_{\abs{P}-1}$ be a path in graph $G'$ with $1 \le \abs{P'} \le k$ and $G'.\glabel(P') \in (\Sigma \setminus \set{\#, \$})^{\ast}$. Then there is a path $P$ in $G$ with $G.\glabel(P) = G'.\glabel(P')$. If $G'$ is a de~Bruijn graph, $G'.\glabel(P)$ is a prefix of $G'.\gkey(v'_{0})$.
\end{lemma}

\begin{proof}
We may assume without loss of generality that graph $G'$ is a de~Bruijn graph.

If $\abs{P} = 1$, every node $v \in G'.\gvalue(v'_{0})$ is a path in graph $G$ with label $X = G'.\glabel(P')$.
Otherwise every edge $(v'_{i}, v'_{i+1}) \in E'$ corresponds to a substring $S_{i}[j, j+k] = X[i] \cdot G'.\gkey(v'_{i+1})$ in the \kcollection{k} $\mathcal{S}$ used for building graph $G'$. Furthermore, the substring has $G'.\gkey(v'_{i})$ as its prefix. We can show by induction from $\abs{P}-2$ to $0$ that $S_{0}[j, j+k]$ has $X$ as its prefix. The sequence of nodes $P = \mathcal{S}.\gnode(S_{0}, j+i) \dotsm \mathcal{S}.\gnode(S_{0}, j+\abs{X}-1)$ is a path in graph $G$ with $G.\glabel(P) = X$.
\end{proof}

@@

Let $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$. By Lemma~\ref{lemma:dbg-fp}, $X$ is a prefix of the substring. We extend the substring to length $\abs{X}+k-1$ without changing its starting position. The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$. If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.

@@

\begin{lemma}[Consequences of maximal pruning]\label{lemma:dbg-maximal}
Let $G' = (V', E')$ be a maximally pruned \orderk{k} de~Bruijn graph. Then
(a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$;
(b) $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all nodes $v' \in V'$, characters $c \in \Sigma$, and nodes $u' \in G'.\gpred(v', c)$; and
(c) key $G'.\gkey(v')$ prefix-matches pattern $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ for all nodes $v' \in G'.\find(X)$.
\end{lemma}

@@

\subsection{GCSA for path indexes}

@@

>>>>

% Shorthands
\newcommand{\kmer}[1]{$#1$\nobreakdash-mer}
\newcommand{\kcollection}[1]{$#1$\nobreakdash-collection}
\newcommand{\kequivalent}[1]{$#1$\nobreakdash-equivalent}
\newcommand{\orderk}[1]{order\nobreakdash-$#1$}
\newcommand{\LFmapping}{LF\nobreakdash-mapping}
\newcommand{\FMindex}{FM\nobreakdash-index}
\newcommand{\patternset}{\ensuremath{(\Sigma \setminus \set{\#, \$})^{\ast}}}

@@

@@

The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} is a permutation of the text with the same combinatorial structure as the suffix array. Given text $T[0, n-1]$, its Burrows-Wheeler transform is a string $\BWT[0, n-1]$, where $\BWT[i] = T[(\SA[i]-1) \bmod n]$. Given the \emph{lexicographic rank} $i$ of suffix $T[\SA[i], n-1]$, we can use \emph{\LFmapping} on the BWT to find the lexicographic rank of the previous suffix $T[(\SA[i]-1) \bmod n, n-1]$. Let
$$
\LF(i) = \Carray[\BWT[i]] + \BWT.\rank(i, \BWT[i]),
$$
where $\Carray[c]$ is the number of occurrences of characters $c' < c$ in the BWT. Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$. We can also generalize the definition for any character $c \in \Sigma$:
$$
\LF(i, c) = \Carray[c] + \BWT.\rank(i, c).
$$
Let $X$ be a string. If there are $i$ suffixes $S'$ of text $T$ such that $S' < X$ in lexicographic order, then there are $\LF(i, c)$ suffixes smaller than string $cX$ in lexicographic order.

@@

\begin{lemma}[No false negatives]\label{lemma:pg-fn}
Let $G' = (V', E')$ be a path graph of graph $G = (V, E)$, and let $X \in \patternset$ be a pattern string. Set $G'.\locate(X)$ contains the start nodes of all paths $P$ in graph $G$ with $G.\glabel(P) = X$.
\end{lemma}

@@

\begin{definition}[Equivalent path graphs]
Let $G$ and $G'$ be two path graphs, and let $k > 0$ be a parameter value. We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in \patternset$ with $1 \le \abs{X} \le k$.
\end{definition}

@@

\begin{lemma}[No short false positives]\label{lemma:dbg-fp}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$, and let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$. Then
(a) $G'.\locate(X)$ is a set of start nodes $v_{0} \in V$ of paths $P$ matching the pattern in graph $G$; and
(b) if $G'$ is a de~Bruijn graph, $X$ is prefix of $G'.\gkey(v'_{0})$ for all $v'_{0} \in G'.\find(X)$.
\end{lemma}

\begin{proof}
If (a) holds for de~Bruijn graphs, it also holds for pruned de~Bruijn graphs. From now on, let $G'$ be a de~Bruijn graph, and let $\mathcal(S)$ be the \kcollection{k} used for building graph $G'$.

(b) If $\abs{X} = 1$, the statement is true by definition. Now let $\abs{X} \ge 2$, and let $P' = v'_{0} \dotsm v'_{\abs{X}}$ be a path matching $X$ in graph $G'$. Every edge $(v'_{i}, v'_{i+1}) \in E'$ corresponds to a substring $S_{i}[j, j+k] = X[i] \cdot G'.\gkey(v'_{i+1})$ in $\mathcal{S}$, and the substring has $G'.\gkey(v'_{i})$ as a prefix. By induction from $\abs{X}-2$ to $0$, substring $S_{i}[j, j+k]$ has $X[i, \abs{X}-1]$ as a prefix.

(a) Now let $S[i, i+k-1] = G'.\gkey(v'_{0})$ be a substring of $\mathcal{S}$, where $v'_{0} \in G'.\find(X)$. Every node $v \in G'.\gvalue(v'_{0})$ is the node $\mathcal{S}.\gnode(S, i)$ for one of such substrings. By (b), $S[i, i+\abs{X}-1] = X$, which corresponds to a path $P$ matching $X$ in graph $G$, with $\mathcal{S}.\gnode(S, i)$ as the start node.
\end{proof}

@@

Let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$. By Lemma~\ref{lemma:dbg-fp}, $X$ is a prefix of the substring. We extend the substring to length $\abs{X}+k-1$ without changing its starting position. The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$. If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.

@@

\begin{lemma}[Consequences of maximal pruning]\label{lemma:dbg-maximal}
Let $G' = (V', E')$ be a maximally pruned \orderk{k} de~Bruijn graph. Then
(a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$;
(b) $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all nodes $v' \in V'$, characters $c \in \Sigma$, and nodes $u' \in G'.\gpred(v', c)$; and
(c) key $G'.\gkey(v')$ prefix-matches pattern $X \in \patternset$ for all nodes $v' \in G'.\find(X)$.
\end{lemma}

@@

\subsection{GCSA for path graphs}

Path graphs are essentially equivalent to the prefix-sorted automata, which were used in the original paper on GCSA \cite{Siren2014}. As our technical details are different from the ones used in that paper, we now describe how the GCSA of a path graph works.

% FIXME refer to the figure in Section 2.4
Let $G = (V, E)$ be a path graph. We sort the nodes $V$ by their keys in lexicographic order and then process the nodes in that order to generate the sequences $\BWT$, $\bvIN$, and $\bvOUT$. For each node $v \in V$, we append
\begin{itemize}
\item $\BWT$ with the predecessor labels $G.\glabel(u)$ for all edges $(u, v) \in E$;
\item $\bvIN$ with the indegree $x = G.\gindegree(v)$ encoded as a binary sequence $0^{x-1} 1$; and
\item $\bvOUT$ with the outdegree $x = G.\goutdegree(v)$ encoded as a binary sequence $0^{x-1} 1$.
\end{itemize}
If node $v \in V$ has lexicographic rank $i$, the range of incoming edges $(u, v) \in E$ to the node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$, with $sp_{in} = 0$ if $i = 0$. The labels of predecessor nodes $u$ are encoded in $\BWT[sp_{in}, ep_{in}]$. Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u, v)$, is equivalent to sorting them by strings $G.\glabel(u) \cdot G.\gkey(v)$. As multiple edges may have the same sort keys, our sorting algorithm must be stable. We can achieve the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.

Consider now the outgoing edges. If the lexicographic rank of node $u \in V$ is $i'$, the range of outgoing edges $(u, v) \in E$ is $[sp_{out}, ep_{out}] = [\bvOUT.\select(i', 1) + 1, \bvOUT.\select(i'+1, 1)]$, with $sp_{out} = 0$ if $i' = 0$. The edges are already sorted by keys $G.\gkey(u)$. All outgoing edges from the same node have the same sort key, as there is nothing to distinguish them from one another. Because graph $G$ is a path graph, we know that key $G.\gkey(u)$ prefix-matches string $G.\glabel(u) \cdot G.\gkey(v)$. The sorting orders are therefore compatible. For every $j \in [sp_{in}, ep_{in}]$ for node $v \in V$, having $\LF(j) \in [sp_{out}, ep_{out}]$ for node $u \in V$ implies an edge $(u, v) \in E$. If we sort the incoming edges $(u, v) \in E$ stably by strings $G.\glabel(u) \cdot G.\gkey(v)$, we have $(u, v) < (u', v')$ in that order, if $G.\gkey(u) < G.\gkey(u')$.

We implement query $G.\find(X)$ with backward searching. Let $X \in \patternset$ be a pattern. If $\abs{X} = 0$, query $G.\find(X)$ returns the lexicographic range $[0, \abs{V}-1]$ containing all nodes of the path graph. Now assume that $\abs{X} \ge 1$ and that query $G.\find(X[i+1, \abs{X}-1])$ has returned range $[sp_{i+1}, ep_{i+1}]$. We want to find the lexicographic range corresponding to the union of sets $G.\gpred(v, X[i])$ over nodes $v \in G.\find(X[i+1, \abs{X}-1])$. We start by mapping the node range $[sp_{i+1}, ep_{i+1}]$ into range
$$
[sp_{in}, ep_{in}] = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]
$$
of incoming edges (with $sp_{in} = 0$ if $sp_{i+1} = 0$). Then we map that range into the corresponding range of outgoing edges
$$
[sp_{out}, ep_{out}] = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, out) - 1].
$$
The lexicographic range containing nodes $G.\find(X[i, \abs{X}-1])$ is
$$
[sp_{i}, ep_{i}] = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
$$

% FIXME from here
Samples
- sampling conditions
- sample encoding
- locate()

Two encodings: general and the one for $\abs{G.\gpred(v, c)} \le 1$. Difference: no general LF in the latter one, must go forward for samples.

@@

These are both based on ST, which requires a maximally pruned DBG

count()
- SadaCount \cite{Sadakane2007a}
- compressible \cite{Gagie2015}

parent()
- CST-NPR: \cite{Fischer2009a}
- CST MEMs: \cite{Ohlebusch2010a}
- used for MEM mapping: \cite{Li2013}
